Design and Implementation of Probabilistic Programming Language Anglican

### 2016-07-31 paper/paper.tex
in progress

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\textit{checkpoints} --- Clojure records of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The records contain fields
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called.

@@

\textit{Memoization} is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation that
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state.

@@

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object
implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
with two methods: \texttt{sample} and \texttt{log-prob}.
The \texttt{sample} \textit{method} returns a random sample and
roughly corresponds to the \texttt{sample} checkpoint,
the \texttt{log-prob} \textit{method} returns the log probability
of the value and roughly corresponds to the \texttt{observe}
checkpoint. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample} or
the \texttt{log-prob} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or re-used.

>>>>

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\textbf{checkpoints} --- Clojure records of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The records contain fields
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called: 
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist val) 'cont)
(->observe dist val cont $state)
\end{lstlisting}

@@

\textbf{Memoization} is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in continuation
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object
implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
with two methods: \texttt{sample*} and \texttt{observe*}.
The \texttt{sample*} \textit{method} returns a random sample and
roughly corresponds to the \texttt{sample} checkpoint,
the \texttt{observe*} \textit{method} returns the log probability
of the value and roughly corresponds to the \texttt{observe}
checkpoint. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample*} or
the \texttt{observe*} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or re-used.
